---
title: 'Assignment 2: Multiple Regression Analysis'
author: "Keith G. Williams - 800690755"
date: "Friday, May 29, 2015"
output: pdf_document
---

Completed as part of DSBA 6201, SUM I 2015.

## The Data Set

The "Boston Housing" dataset recorded properties of 506 housing zones in the Greater Boston area. Typically, one is interested in predicting `MEDV` (median home value) based on other attributes.

Here is a list of attribute information:
1. `CRIM`: per capita crime rate by town
2. `ZN`: proportion of residential land zoned for lots over 25,000 $ft^2$
3. `INDUS`: proportion of non-retail business acres per town
4. `CHAS`: Charles River dummy variable (=1 if tract bounds river; 0 otherwise)
5. `NOX`: nitric oxides concentraion (parts per 10 million)
6. `RM`: average number of rooms per dwelling
7. `AGE`: proportion of owner-occupied units built prior to 1940
8. `DIS`: weighted distances to five Boston employment centers
9. `RAD`: index of accessibility to radial highways
10. `TAX`: full-value property-tax rate per $10,000
11. `PTRATIO`: pupil-teacher ratio by town
12. `LSTAT`: % lower status of the population
13. `MEDV`: median value of owneer-occupied homes in $1000's

```{r}
file <- "~/DSBA 6201/Boston Housing.csv"
bos <- read.csv(file)
```

## Pre-processing

`MEDV` has somewhat longish tail and is not so normally distributed, so we will take the log transform, and then predict `LMEDV` instead.

```{r}
library(ggplot2)

# plot histogram of median value
h <- ggplot(bos, aes(MEDV)) + geom_histogram()
h
```

```{r}
bos$LMEDV <- log(bos$MEDV)
l <- ggplot(bos, aes(LMEDV)) + geom_histogram()
l
```

## Questions

1. Please perform the multicollinearity diagnosis based on the VIF calculation results. Do we need to drop any variables that might have multicollinearity concerns?

```{r}
library(car)
fit <- lm(LMEDV ~ . - MEDV, data = bos)
vif(fit)
```

The `TAX` variable has a high variance inflation factor, and can be seen to be highly correlated with `RAD`, so it will be removed from the model.

```{r}
cor(bos$TAX, bos$RAD)
fit2 <- update(fit, . ~ . - TAX)
vif(fit2)
```

2. Please run the linear regrssion analyses.  
- Use the stepwise model selection approach to determine the final model. Drop variables based on thier significance.  
- Report summary output for *each step*, including ANOVA, $R^2$, and parameter estimates.

```{r}
step(fit2, direction = "backward")
```

According to the backward step model selection, the final model should include `CRIM`, `CHAS`, `NOX`, `RM`, `DIS`, `RAD`, `PTRATIO`, and `LSTAT`.

```{r}
mod1 <- update(fit2, . ~ . - AGE)
mod2 <- update(mod1, . ~ . - INDUS)
mod3 <- update(mod2, . ~ . - ZN)

# analysis of variance
anova(fit2, mod1, mod2, mod3)
```

```{r}
fit3 <- lm(LMEDV ~ CRIM + CHAS + NOX + RM + DIS + RAD + PTRATIO + LSTAT, data = bos)
```

3. How do you interpret the final model?
```{r}
summary(fit3)
```

The natural log of median home value ($1000) changes by the estimated coefficient for each feature for each unit change in that feature, while all other features are held constant. To use `CRIM` as an example:

$log(MEDV)$ decreases by $0.010619 \pm 0.0013399$ for every 1 unit increase in the per capita crime rate for the same `CHAS`, `NOX`, `RM`, `DIS`, `RAD`, `PTRATIO`, and `LSTAT`.

The 95% confidence interval for each coefficient is calculated by obtaining the 0.975 t-quantile on 497 degrees of freedom multiplied by the standard error for that coefficient. This value is added and subtracted from the fitted value at each $X_i$ to get the upper and lower bounds.

Finally the adjusted $R^2$ value of 0.7706 means that 77% of the variance in the data are explained by the model.